$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-05-01 paper/appendix.tex
Keep the contexts subsection short

\todo{we haven't introduced the forall and implicit syntax}
\begin{nidefinition}
  The algebra on multiplicities must be \emph{partial}, \emph{functional}, \emph{cancelative}, \emph{associtive}, and \emph{commutative}, and needs a \emph{minimal neutral element}.
  It also requires a special element $\one$ that is to be used to count inputs and outputs.
  We define the laws of such a \emph{partial commutative monoid} as a type $\Quantifier_C$ on a carrier set $C$ in \autoref{fig:multiplicities} --- we use a terniary relation to model the monoidal operation.

  \begin{figure}[h]
  \begin{equation}
  \begin{aligned}
    &\zero                  &:{} &                      &        & C \\
    &\one                   &:{} &                      &        & C \\
    &\op{\_}{\_}{\_}        &:{} &                      &        & C \to C \to C \to et \\
    &\field{\cdot-compute}  &:{} &\forall y z           & \to \; & \type{DEC} \; (\type{\exists} x . \; (\op{x}{y}{z})) \\
    &\field{\cdot-unique}   &:{} &\forall \{x x' y z\}  & \to \; & \op{x'}{y}{z} \to \op{x}{y}{z} \to x' \equiv x \\
    &\field{\cdot-unique^l} &:{} &\forall \{x y y' z\}  & \to \; & \op{x}{y'}{z} \to \op{x}{y}{z} \to y' \equiv y \\
    &\field{0\cdot-min^l}   &:{} &\forall \{y z\}       & \to \; & \op{\zero}{y}{z} \to y \equiv \zero \\
    &\field{\cdot-id^l}     &:{} &\forall \{x\}         & \to \; & \op{x}{\zero}{x} \\
    &\field{\cdot-comm}     &:{} &\forall \{x y z\}     & \to \; & \op{x}{y}{z} \to \op{x}{z}{y} \\
    &\field{\cdot-assoc}    &:{} &\forall \{x y z u v\} & \to \; & \op{x}{y}{z} \to \op{y}{u}{v} \to \type{\exists} w . \; (\op{x}{u}{w} \times \op{w}{v}{z})
  \end{aligned}
  \end{equation}
  \caption{Quantifier algebra $\Quantifier_C$ algebra on a partial commutative monoid.}
  \label{fig:multiplicities}
  \end{figure}
\end{nidefinition}

>>>>

\todo{we haven't introduced the forall and implicit syntax}
\todo{introduce DEC and exists}

Our usage algebra is a ternary relation $\op{x}{y}{z}$ that is \emph{partial}, \emph{functional}, \emph{cancelative}, \emph{associtive}, and \emph{commutative}.
It has a \emph{neutral element} $\zero$ that is absorbed on either side, and that is also \emph{minimal}.
It has an element $\one$ that is used to count inputs and outputs.
These algebraic laws are defined as a record $\Quantifier_C$ on a carrier set $C$ in \autoref{fig:multiplicities}.
We use a terniary relation to model the partial monoidal operation.

\begin{figure}[h]
\begin{equation}
\begin{aligned}
  &\zero                  &:{} &                      &        & C \\
  &\one                   &:{} &                      &        & C \\
  &\op{\_}{\_}{\_}        &:{} &                      &        & C \to C \to C \to et \\
  &\field{\cdot-compute}  &:{} &\forall y z           & \to \; & \type{DEC} \; (\type{\exists} x . \; (\op{x}{y}{z})) \\
  &\field{\cdot-unique}   &:{} &\forall \{x x' y z\}  & \to \; & \op{x'}{y}{z} \to \op{x}{y}{z} \to x' \equiv x \\
  &\field{\cdot-unique^l} &:{} &\forall \{x y y' z\}  & \to \; & \op{x}{y'}{z} \to \op{x}{y}{z} \to y' \equiv y \\
  &\field{0\cdot-min^l}   &:{} &\forall \{y z\}       & \to \; & \op{\zero}{y}{z} \to y \equiv \zero \\
  &\field{\cdot-id^l}     &:{} &\forall \{x\}         & \to \; & \op{x}{\zero}{x} \\
  &\field{\cdot-comm}     &:{} &\forall \{x y z\}     & \to \; & \op{x}{y}{z} \to \op{x}{z}{y} \\
  &\field{\cdot-assoc}    &:{} &\forall \{x y z u v\} & \to \; & \op{x}{y}{z} \to \op{y}{u}{v} \to \type{\exists} w . \; (\op{x}{u}{w} \times \op{w}{v}{z})
\end{aligned}
\end{equation}
\caption{Quantifier algebra $\Quantifier_C$ algebra on a partial commutative monoid.}
\label{fig:multiplicities}
\end{figure}

\section{Indexed Set of Usage Algebras}
\label{app:indexed-set-usage-algebra}

\begin{figure}[h]
  \begin{equation}
    \begin{split}
      &\field{IDX}          &:{} &\Set \\
      &\field{\exists IDX}  &:{} &\field{IDX} \\
      &\field{CARRIER}      &:{} &\field{IDX} \to \Set \\
      &\field{QUANTIFIERS}  &:{} &(i : \field{IDX}) \to \Quantifier_{\field{CARRIER}_i}
    \end{split}
  \end{equation}
  \caption{Indexed set of usage algebras.}
  \label{fig:indexed-multiplicities}
\end{figure}

### 2020-05-01 paper/paper.tex
Keep the contexts subsection short

\subsection{Contributions}
Our contributions are the following:
\begin{itemize}
\item \textbf{Formalisation of the syntax and the semantics of the \picalc{}}:
  \autoref{syntax} uses type level de Bruijn indices \cite{} to introduce a syntax that is well-scoped by construction: every free variable is accounted for in the type of the process that uses it.
  \autoref{semantics} provides an operational semantics for the untyped \picalc{}, prior to any typing.
  This operational semantics is given as a reduction relation on processes.
  The reduction relation tracks at the type level what channel the communication occurs on, so that this information can then be used to state subject reduction --- ak-a type preservation.
  It is also defined modulo structural congruence --- a relation that acts as a quotient type that gets rid of the unecessary syntactic minutiae introduced by the syntax of the \picalc{}.

@@

Scope restriction $(\new{}x) \; P$ and input $\recvp{x}{y}. \; P$ both introduce bound names in $P$.
In order to mechanise the \picalc{} syntax in Agda, we need to deal with bound names in continuation processes.
Names are cumbersome to mechanise: inserting a new variable into a context means proving that its name differs from all other variable names in context.
Furthermore, names do not guarantee that the syntax well-scoped.
Instead, we use de Bruijn indices \cite{}, where a natural number $n$ (aka \emph{index}) is used to refer to the variable introduced $n$ binders ago --- hence binders no longer introduce names.
\begin{example}
  \begin{equation*}
    \begin{split}
      & (\new{}x ) && (\comp {\recvp{x}{y} && . \; \sendp{y}{\texttt{"Hello world!"}} . \; \PO} {(\new{} y) && (\sendp{x}{y} . \; \recvp{y}{z} && . \; \PO)})
      & \text{names}
      \\
      & \new{} && (\comp {\recv{0} && . \; \send{0}{\texttt{"Hello world!"}} . \; \PO} {\new{} && (\send{1}{0} . \; \recv{0} && . \; \PO)})
      & \text{de Bruijn}
      \\
    \end{split}
  \end{equation*}
\end{example}
That is, terms at different \emph{depths} use different indices to refer to the same binding.

@@

\subsection{Typing Contexts}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\label{contexts}
In a resource-aware \picalc{}, usage annotations are transformed as a result of input and output; types however are preserved.
In our case, usage annotations do not depend on types, and keeping types and usages in separate contexts allows for a more modular approach in our proofs.

\begin{nidefinition}
  We define types (\autoref{fig:prectx}) as either unit types, base types or channel types.
  The unit type $\unit$ serves as a proof of inhabitance for types (we can trivially obtain it), and is used to such end in subject congruence --- for the $\new \PO_{\suc n} \eqeq \PO_n$ rule.
  The base type $\base{n}$ uses a natural number $n$ to represent types.
  This natural number can then be interpreted as a type in the host language.
  Using natural numbers instead of type universes is convenient: less universe polymorphism poluting specifications.

  The channel type $\channel{t}{x}$ uses $t$ to stand for the type of data that is being transmitted over the channel, and $x$ to stand for what usages of that data are transmitted.

  We keep track of types in a length-indexed \emph{pre-context} $\PreCtx_n$.

@@

\begin{nidefinition}
  We use indexed sets of usage algebras to allow several usage algebras to coexist in our type system.
  An indexed set of usage algebras is a set of indices such that for every index there exists a carrier $C$ with a $\Quantifier_C$ algebra defined on it, as captured by \autoref{fig:indexed-multiplicities}.
  We require the index type to be inhabited -- proving subject congruence on $\new \PO_{\suc n} \eqeq \PO_n$ requires an inhabited carrier set.

  \begin{figure}[h]
  \begin{equation}
  \begin{split}
    &\field{IDX}          &:{} &\Set \\
    &\field{\exists IDX}  &:{} &\field{IDX} \\
    &\field{CARRIER}      &:{} &\field{IDX} \to \Set \\
    &\field{QUANTIFIERS}  &:{} &(i : \field{IDX}) \to \Quantifier_{\field{CARRIER}_i}
  \end{split}
  \end{equation}
  \caption{Indexed set of partial commutative monoids.}
  \label{fig:indexed-multiplicities}
  \end{figure}
\end{nidefinition}

\begin{nidefinition}
  A context of usages $\Ctx_{idxs}$ is indexed by a length-indexed context of indices $\Idxs_n$, as shown in \autoref{fig:ctx}.
  \begin{figure}[h]
  \begin{mathpar}
    \datatype
    {n : \N}
    {\Idxs_n : \Set}
    \; \textsc{Idxs}

@@

\begin{note}
  We lift the monoidal operation on usage annotations $\opsquared{x}{y}{z}$ to a monoidal operation $\opctx{\Gamma}{\Delta}{\Xi}$ on usage contexts with a common underlying context of indices.
  All the properties of the partial monoid are trivially lifted.
\end{note}

@@

We make use \emph{leftover typing} \cite{}, an approach that, in addition to the usual typing context and input usage context, adds an extra output usage context to the typing rules.
This output context contains the \emph{leftovers} of the process being typed --- its unused multiplicities.
Crucially, the leftovers can then be used as an input to some other typing derivation.
Adopting this techinque has three main advantages:
\begin{itemize}
  \item
    The input and output actions no longer transform the usage context through functions.
    Instead, they use a variable reference type with both an input and an output usage context that encapsulates these transformations.
  \item
    Parallel composition no longer needs an explicit context splitting proof --- a proof $\opctx{\Gamma_m}{\Gamma_l}{\Gamma_r}$ where $\Gamma_l \vdash P$, $\Gamma_r \vdash Q$ and $\Gamma_m \vdash \comp{P}{Q}$.
    Such a proof could be obtained through automated search --- but it is better yet to avoid needing it.
    With leftover typing, we use the leftovers of $P$ to type $Q$.
  \item
    Theorems like framing ( \autoref{framing}), weakening ( \autoref{weakening}) and strengthening ( \autoref{strengthening}) become suddenly stateable.
\end{itemize}

>>>>

\subsection{Contributions}
Our contributions are the following:
\begin{itemize}
\item \textbf{Formalisation of the syntax and the semantics of the \picalc{}}:
  \autoref{syntax} uses type level de Bruijn indices \cite{deBruijn1972} to introduce a syntax that is well-scoped by construction: every free variable is accounted for in the type of the process that uses it.
  \autoref{semantics} provides an operational semantics for the untyped \picalc{}, prior to any typing.
  This operational semantics is given as a reduction relation on processes.
  The reduction relation tracks at the type level what channel the communication occurs on, so that this information can then be used to state subject reduction --- ak-a type preservation.
  It is also defined modulo structural congruence --- a relation that acts as a quotient type that gets rid of the unecessary syntactic minutiae introduced by the syntax of the \picalc{}.

@@

Scope restriction $(\new{}x) \; P$ and input $\recvp{x}{y}. \; P$ both introduce bound names in $P$.
In order to mechanise the \picalc{} syntax in Agda, we need to deal with bound names in continuation processes.
Names are cumbersome to mechanise: inserting a new variable into a context means proving that its name differs from all other variable names in context.
Furthermore, names do not guarantee that the syntax well-scoped.
Instead, we use de Bruijn indices \cite{deBruijn1972}, where a natural number $n$ (aka \emph{index}) is used to refer to the variable introduced $n$ binders ago --- hence binders no longer introduce names.
\begin{example}
  \begin{equation*}
    \begin{split}
      & (\new{}x ) && (\comp {\recvp{x}{y} && . \; \sendp{y}{\texttt{"Hello world!"}} . \; \PO} {(\new{} y) && (\sendp{x}{y} . \; \recvp{y}{z} && . \; \PO)})
      & \text{names}
      \\
      & \new{} && (\comp {\recv{0} && . \; \send{0}{\texttt{"Hello world!"}} . \; \PO} {\new{} && (\send{1}{0} . \; \recv{0} && . \; \PO)})
      & \text{de Bruijn}
      \\
    \end{split}
  \end{equation*}
\end{example}
That is, terms at different \emph{depths} use different indices to refer to the same binding.

@@

\subsection{Typing Contexts}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\label{contexts}

We keep typing contexts and usage contexts separate.
The former are preserved throughout a typing derivation; the latter are transformed as a result of input, output, and context splits.

\begin{nidefinition}
  \autoref{fig:prectx} defines types as either unit types, base types or channel types.
  The unit type $\unit$ serves as a proof of inhabitance for types.
  The base type $\base{n}$ uses natural numbers as placeholders for types --- the host language can interpret the former into the latter.
  This is for convenience, so that there is less universe polymorphism poluting specifications.
  The type $\channel{t}{x}$ of a channel determines what type of data and what usage annotations are sent over that channel.
  We keep track of types in a length-indexed \emph{pre-context} $\PreCtx_n$.

@@

\begin{nidefinition}
  We use indexed sets of usage algebras to allow several usage algebras to coexist in our type system.
  An indexed set of usage algebras is a set of indices together with an interpretation of indices into types, and an interpretation of indices into usage algebras on that type.
  This indexed set must be non-empty.
  Refer to \autoref{fig:indexed-multiplicities} in \autoref{app:indexed-set-usage-algebra} for a formal definition.
\end{nidefinition}

\begin{nidefinition}
  We keep track of usages annotations in a context $\Ctx_{idxs}$ that is indexed by a length-indexed context of indices $\Idxs_n$, as shown in \autoref{fig:ctx}.

  \begin{figure}[h]
  \begin{mathpar}
    \datatype
    {n : \N}
    {\Idxs_n : \Set}
    \; \textsc{Idxs}

@@

\begin{note}
  We lift the monoidal operation on usage annotations $\opsquared{x}{y}{z}$ to a monoidal operation $\opctx{\Gamma}{\Delta}{\Xi}$ on usage contexts that have a common underlying context of indices.
  All the properties of the partial monoid are trivially lifted.
\end{note}

@@

We make use \emph{leftover typing} \cite{}, an approach that, in addition to the usual typing context and input usage context, adds an extra output usage context to the typing rules.
This output context contains the \emph{leftovers} of the process being typed --- its unused multiplicities.
Crucially, the leftovers can then be used as an input to some other typing derivation.
Adopting this techinque has three main advantages:
\begin{itemize}
  \item
    The input and output actions no longer transform the usage context through functions.
    Instead, they use a variable reference type with both an input and an output usage context that encapsulates these transformations.
  \item
    Parallel composition no longer needs an explicit context splitting proof --- a proof $\opctx{\Gamma_m}{\Gamma_l}{\Gamma_r}$ where $\Gamma_l \vdash P$, $\Gamma_r \vdash Q$ and $\Gamma_m \vdash \comp{P}{Q}$.
    Such a proof could be obtained through automated search --- but it is better yet not to need it.
    With leftover typing, we use the leftovers of $P$ to type $Q$.
  \item
    Theorems like framing ( \autoref{framing}), weakening ( \autoref{weakening}) and strengthening ( \autoref{strengthening}) become suddenly stateable.
\end{itemize}
